---
title: "Getting started with HLCtools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with HLCtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`HLCtools` calculates Herd Lying Concordance (HLC) metrics from individual animal lying-behaviour data.

HLC is a framework for quantifying group-level behavioural cohesion. Instead of classifying an interval as synchronous using a fixed threshold, HLC uses the distribution of individual animal lying behaviour within each interval.

The package currently supports four HLC implementations:

- `HLC_SD`: standard-deviation-based HLC
- `HLC_MAD`: mean-absolute-deviation-based HLC
- `HLC_IQR`: interquartile-range-based HLC
- `HLC_ENT`: entropy-based HLC

The package also calculates lying-weighted HLC, which combines group cohesion with the proportion of the interval spent lying.

# Load package and example data

```{r}
library(HLCtools)

data(example_lies)

example_lies
```

# Required data format

The input data should contain one row per animal per time interval.

At minimum, the data should contain:

- a group identifier;
- an animal identifier;
- a time interval identifier;
- lying time within the interval.

Optional but usually useful columns include:

- day or date;
- week or experimental period.

In `example_lies`, the relevant columns are:

```{r}
str(example_lies)
```

# Calculate interval-level HLC

```{r}
hlc_intervals <- calculate_hlc(
  data = example_lies,
  group = group,
  animal = cow,
  day = day,
  period = week,
  interval = time,
  lying = lying,
  interval_min = 15,
  methods = c("sd", "mad", "iqr", "entropy"),
  sync_thresholds = c(0.6, 0.7, 0.8, 0.9),
  add_lying_weighted = TRUE
)

hlc_intervals
```

The output contains one row per group-time interval.

Important columns include:

- `mean_lying`: mean lying minutes in the interval;
- `lying_prop`: mean lying proportion in the interval;
- `HLC_SD`, `HLC_MAD`, `HLC_IQR`, `HLC_ENT`: unweighted HLC metrics;
- `HLC_SD_LYING`, `HLC_MAD_LYING`, `HLC_IQR_LYING`, `HLC_ENT_LYING`: lying-weighted HLC metrics;
- `sync60`, `sync70`, `sync80`, `sync90`: threshold synchrony indicators.

# Summarise HLC by day

```{r}
hlc_daily <- summarise_hlc_daily(
  data = hlc_intervals,
  group = group,
  day = day,
  period = week,
  interval_min = 15
)

hlc_daily
```

The daily summary contains one row per group-day.

Daily HLC columns are means of interval-level HLC values. Synchrony threshold columns are summarised as minutes per day.

# Rank HLC methods for a dataset

Different experiments may favour different HLC implementations. For example, one dataset may favour SD-based HLC, whereas another may favour MAD-, IQR-, or entropy-based HLC.

`rank_hlc_methods()` provides a descriptive ranking of available HLC implementations.

```{r}
rank_hlc_methods(
  data = hlc_daily,
  group = group,
  period = week,
  lying_prop = mean_lying_prop
)
```

The ranking does not prove that one method is universally best. It provides a transparent dataset-specific summary to support method selection.


# Interpretation

Unweighted HLC describes group-level behavioural cohesion.

A high unweighted HLC value means that animals behaved similarly within the interval. This can occur when animals are uniformly lying or uniformly standing.

Lying-weighted HLC describes cohesive lying specifically.

A high lying-weighted HLC value means that animals were both behaviourally cohesive and lying.

# Using HLC and lying-weighted HLC together

The two metrics can be interpreted together:

- high HLC and high lying-weighted HLC: cohesive collective lying;
- high HLC and low lying-weighted HLC: cohesive behaviour, but not mainly lying;
- low HLC: fragmented group behaviour.

# Available HLC methods

```{r}
hlc_methods()
```

# Notes on method choice

`HLCtools` does not assume that one dispersion basis is universally best.

The choice of SD, MAD, IQR, entropy, or another implementation should depend on:

- the biological question;
- group size;
- sampling interval;
- behaviour type;
- data quality;
- intended monitoring application.

Researchers are encouraged to calculate multiple HLC variants and report the dispersion basis used.
